---
title: "HIPAA Breach Analysis"
author: "Michael Molloy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    df_print: kable
    includes:
          after_body: "./assets/html/footer.html"
    
---


```{r include = FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(stringdist)
library(reshape2)
library(stringr)
library(wordcloud2)
library(tm)
library(usmap)
library(lubridate)
library(tidyverse)


# Define multmerge function - merges multiple csv files together
multmerge = function(mypath) {
  filenames=list.files(path=mypath, full.names = TRUE)
  datalist = lapply(filenames, function(x){
    read.csv(file=x,header=T,stringsAsFactors = F)
  })
  Reduce(function(x,y){
    merge(x,y,all = TRUE)
  }, datalist)
}
# Load files
breaches = multmerge("./data/source/")
entity_name_map = read.csv("./data/project/entity_name_map.csv", header = TRUE,sep = ",", stringsAsFactors = FALSE)

# clean up the dates
breaches <- breaches %>% 
 mutate(date = mdy(breaches$Breach.Submission.Date),
         year = year(date),
         month = floor_date(date, unit = "month"))
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
individuals_by_month <- breaches %>% 
   group_by(year,month) %>% 
  summarize(total_records = sum(na.omit(Individuals.Affected)))

count_by_month <- breaches %>% 
    group_by(year,month) %>% 
  summarize(count = n())

breaches_by_month <- merge (individuals_by_month, count_by_month)

breaches_by_month %>%
  ggplot(aes(month, count, size = total_records)) +
    geom_point(color = "purple") +
   scale_x_date(NULL, date_labels = "%Y", breaks = "year") +
    geom_smooth(method = "lm", show.legend= FALSE) +
  scale_size_continuous(breaks = c(20000000,40000000,60000000),labels = c("20M","40M","60M")) +
  labs(title = "Number of Breaches by Month", y = "# of breaches", x = "Year", size = "# of records") 
```

# Types of Breaches
The most common types of breaches are hacking, theft, and unauthorized access. Breach reports can contain multiple breach types, so some parsing had to be done in order to create a meaningful chart on this field.  The graph below shows the number of reported breaches per month distinguished by their breach type using color.  The size of the dot, similar to our previous chart, shows the relative sum of the records (i.e. individuals affected) for the breaches in the month.  

A few things stand out in this chart. Most notably that the reporting of theft has been declining sharply and the reports of unauthorized access and hacking have been increasing. My hypothesis on why the reports of theft have dropped is that the prevalence of encrypted storage has increased.  This is a purely anecdotal observation from my experience in IT Risk over the past 10 years, but it makes sense. If a laptop is stolen but its hard drive is encrypted, OCR does not consider the data on that hard drive accessible to an unauthorized user and therefore would not constitute a reportable breach. 

My other observation is that the biggest breaches (i.e. ones that affect the most individuals) are related to hacking. When you combine that with the increase in hacking incidents it paints a dire picture for individual privacy. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

hack_by_month <- breaches %>% 
  filter(str_detect(Type.of.Breach, "Hack")) %>% 
  group_by(year,month) %>% 
  summarize(count = n(), records = sum(Individuals.Affected)) %>% 
  mutate(type = "Hack")

theft_by_month <- breaches %>% 
  filter(str_detect(Type.of.Breach, "Theft")) %>% 
  group_by(year,month) %>% 
  summarize(count = n(), records = sum(Individuals.Affected)) %>% 
  mutate(type = "Theft")

access_by_month <- breaches %>% 
  filter(str_detect(Type.of.Breach, "Access")) %>% 
  group_by(year,month) %>% 
  summarize(count = n(), records = sum(Individuals.Affected)) %>% 
  mutate(type = "Unauthorized Access")

loss_by_month <- breaches %>% 
  filter(str_detect(Type.of.Breach, "Loss")) %>% 
  group_by(year,month) %>% 
  summarize(count = n())%>% 
  mutate(type = "Loss")
 
 disposal_by_month <- breaches %>% 
   filter(str_detect(Type.of.Breach, "Improper")) %>% 
   group_by(year,month) %>% 
   summarize(count = n()) %>% 
   mutate(type = "Improper Disposal")
 
 other_by_month <- breaches %>% 
   filter(str_detect(Type.of.Breach, "Other")) %>% 
   group_by(year,month) %>% 
   summarize(count = n()) %>% 
   mutate(type = "Other")

type_by_month <- rbind(hack_by_month,
                       theft_by_month,
                       access_by_month,
                       loss_by_month,
                       disposal_by_month,
                       other_by_month)

type_by_month %>% 
  ggplot(aes(month, count, color = type, size = records)) +
  geom_point(show.legend = TRUE) +
  scale_x_date(breaks = "year", date_labels = "%Y") +
  ylim(0,25) +
  geom_smooth(method="lm", show.legend = FALSE) +
  scale_size_continuous(breaks = c(20000000,40000000,60000000),labels = c("20M","40M","60M")) +
  labs(title = "Types of Data Breaches", color = "Breach Type", x = "Year", y = "# of breaches per month", size = "# of records")
```


# Breach Map {.tabset}
## Total Breaches by State
The map below shows the total number of breaches by state. At first glance, it appears as though covered entities in California, Texas, and Florida have much higher occurrences of data breaches than covered entities in other states.  However, this map looks a lot like a map of state populations from the 2015 census (see the next tab).
```{r echo=FALSE}
# count numer of breaches by state
dfState <- breaches %>%
  group_by("state" = State) %>%
  summarise("Total" = length(State)) %>%
  arrange(desc(Total))

# plot total breaches on map
plot_usmap(data = dfState, values = "Total",regions = "states", lines = "white") + 
  scale_fill_continuous(name = "Breaches", label = scales::comma, high = "red", low = "white") + 
  theme(legend.position = "right") +
  labs(title = "Breaches by State", subtitle = "From Oct 2009- Apr 2019", caption = "Office of Civil Rights, https://ocrportal.hhs.gov/ocr/breach/breach_report.jsf")
```

## Population Map
There appears to be a correlation between the number of breaches in each state and the population.  It might be worth exploring this further. I'd want to test the hypothesis that a higher population means more covered entities (i.e. hospitals) which are available to be breached in a given state. The next tab shows a more normal distribution of breaches across the US.
```{r echo=FALSE}
plot_usmap(data = statepop, values = "pop_2015", lines = "white") + 
  scale_fill_continuous(name = "Population", label = scales::comma, high = "red", low = "white") + 
  theme(legend.position = "right") +
  labs(title = "Population by State", subtitle = "2015",caption = "United States Census")
```

## Breaches by Population
I normalized the breaches by dividing the number of breaches by the total state population. I expect this represents a more accurate picture, but statistical analysis would need to be done to determine whether a particular state is more likely to have a breach than others.
```{r echo=FALSE}
dfState <- breaches %>%
  group_by("abbr" = State) %>%
  summarise("Breaches" = length(State)) %>%
  arrange(desc(Breaches))

breach_pop <- merge(dfState, statepop, by = "abbr")
breach_pop$Breach.Percent <- as.numeric(breach_pop$Breaches/breach_pop$pop_2015, digits = 2, format = "f")*1000000

plot_usmap(data = breach_pop, values = "Breach.Percent",regions = "states", lines = "white") + 
  scale_fill_continuous(name = "Ratio", label = scales::comma, high = "red", low = "white") + 
  theme(legend.position = "right") +
  labs(title = "Breaches as a Ratio of Population", subtitle = "Breaches from Oct 2009- Apr 2019/ State Populations from 2015 US Census")
```


# Covered Entities {.tabset}
## Most Breached Entities
I also wanted to see which covered entities incurred the most breaches. I found quality issues with the covered entity name column.  I used a clustering algorithm to automatically group like names, then created a "mapping table" (see other tab) of rules based on those clusters. After running the mapping table logic, I was able to count all the breaches by entity and get a more accurate picture of the covered entities with the most reported breaches.

```{r echo=FALSE}
# change covered entity names to a common name
for (row in 1:NROW(entity_name_map)) {
  from <- entity_name_map[row,1]
  to <- entity_name_map[row,2]
  breaches$Name.of.Covered.Entity <- str_replace_all(breaches$Name.of.Covered.Entity,from,to)
}

# Count the number of occurences for each Entity
dfName <- breaches %>%
  group_by("Entity Name" = Name.of.Covered.Entity) %>%
  summarise("Total" = length(Name.of.Covered.Entity)) %>%
  arrange(desc(Total))

head(dfName,20) %>% 
ggplot(aes(reorder(`Entity Name`,Total),Total, fill = as.factor(Total))) +
  geom_col() +
  guides(fill="none")+
  coord_flip() +
  ylim(0,25) +
  labs(title = "Entities with the most breaches", x = "", y = "# of breaches")
```

## Mapping Table

```{r echo=FALSE}
# print table
kable(entity_name_map) %>%
  kable_styling(bootstrap_options = c("striped","hover", "condensed","responsive"),fixed_thead = T) %>%
  column_spec(1, bold = T, border_right = T) %>%
  column_spec(2)
```

# Conclusion
Cyber criminals are attacking organizations of all sizes and industry verticals as IT and Security departments struggle to keep up with the latest threats.  The Healthcare industry has seen an increase in the number of breaches over the past few years, especially those related to hacking.  Understanding the threat landscape is an important part of increasing the privacy protections over patient data and identifying solutions that reduce breaches.  


